Assessing variations in the chemistry of fossil- and modern-*Pinus* pollen

### 2021-06-09 manuscript.Rmd
rearranged R/D; added Jardine 2021 comparison

@@

After visual inspection of the spectra, the chemical differences between modern and fossil pollen were explored with a PLSR model.
The ordination showed that the fossil pollen are clearly separated from the modern pollen (Fig \@ref(fig:ordinations) a).
The regression coefficients reflect our observations from the visual inspection and show high importance of the 1745 lipid peak, the 1500 - 1800 region and the 1170 peak (Fig \@ref(fig:ordinations) b and c).

## Differences between fossil pollen
Using this PLSR model the acetolysed pollen is clearly separated from the non acetolysed pollen, with a much higher distance between samples for acetolysed pollen compared to non acetolysed pollen (Fig. \@ref(fig:ordinations) d).
In addition, there is some differentiation between the cores, samples from Tiefer See are more separate from samples of the other two cores, which show more similar scores in the ordination.

## Anomalies in fossil pollen spectra

@@

The 1090 cm^-1^ peak dissappears after acetolysis from all spectra as does the 1630 cm^-1^ peak in Tiefer See. 
The region 1000 - 1200 cm^-1^ is now dominated in acetolysed fossil pollen by two peaks and its shoulders at 1040 and 1170 cm^-1^.
This 1040 cm^-1^ has high importance for the PLSR model differentiating acetolysed and non acetolysed pollen (Fig. \@ref(fig:ordinations) e and f).

@@

# Discussion

.Overall intro to discussion
Our aim with this study was to explore the use of a new type of detector (FPA-MCT) in combination with a FTIR microscope to explore the chemical variations of fossil pollen.
Our goal was to also to explore the effects of two differemt sample-processing methods on fossil pollen and compare the spectra we obtained with modern samples.

.difference to fresh pollen (FM: have to mention Jardine 21 here)
One of the clearest differences in the results in our study was that fossil pollen spectra (either treated or untreated) show major differences from the fresh- (i.e. modern pollen samples).
As expected, in the fossil-pollen samples (both treated and untreated), lipid and protein peaks disappeared and there are substantial changes in the carbohydrate region in fossil pollen.
From the regression coefficients, we see that the largest changes are in the 1500 to 1750 cm^-1^ region, which have many lipid and protein peaks (eg. 1745 and 1545 cm^-1^ etc).
**Here you need to add in more comparisons of the other published papers-the two Jardine papers and others which compare them**

Within the fossil pollen samples, there were also differences in spectra identified between acetolysised and non-acetolysised pollen grains. 
Acetolysed pollen showed two dominant peaks in the carbohydrate region which are not visible in the modern spectra at 1040 and 1170 - 1190 cm^-1^ or the non acetolysed pollen.
This change in the 1170 - 1190 cm^-1^ region is unique to acetolysized pollen and not visible in non acetolysed pollen and had higher regression coefficients that the surrounding wavenumbers (Fig. \@ref(fig:ordinations)).
The presence of this new peak in the shoulder of the 1170 cm^-1^ peak, which can indicate sporopollenin and carbohydrates, suggest changes in the pollen chemistry.
These changes may be related to removal and or alteration of grain wall carbohydrates (cellulose etc) as well as changes to sporopollenin due to acetolysis. [FM: this would be answered by acetolysis of the fresh grains!!]
In non acetolysed pollen the 1190 peak is not visible in any of the cores, which indicates that it is not due to aging/diageneis, but acetolysis.

.spectral interference
One additional finding we did not expect was that the non acetolysed pollen showed a wide peak covering part of the carbohydrate region at 1040 cm^-1^.
This peak in the non acetolysed samples is most likely biogenic silica, when compared to this biogenic silica spectra from diatoms [@alipourInfraredMicrospectroscopyBionanomaterials2016] and hydrated silica spectra [@SilicaATRIRSpectrum].
During the measurements with the microscope, we applied our extracted pollen as droplets onto ZnSe slides and waited for the water to evaporate.
After the droplets had dried we observed a crystalline layer over the same area that covered all particles and pollen (ref Foto in SM).
From the chemical spectra, it is most likely that that it is dissolved silica that crystallised, although the acetolysis treatment removed this peak, which may be because of the extended washing with several chemicals and water during standard acetolysis.
**I think you need to explain this peak in the context of palaeoecological research a bit- could you talk about the correction factors that might be needed if you were to apply this method everywhere? And that different sediment context might create different baseline spectra so understanding sedimentary context is essential?**

the characteristic sporopollenin peaks are all present, but since so much of the rest of the spectra changes, we cant clearly associate any relative changes in these peaks to changes in composition of the sporopollenin or simply changes that only affect other components of the pollen (e.g. carbohydrates).

We saw large overlap between MFM and DAL in the PLSR ordination , both in non acetolysed and acetolysed pollen.
This seems to indicate that pine pollen from different cores are chemically quite similar, even cores that are far apart.
TSK pollen on the other hand could be clearly separated from the other two cores, in both SPT and acetolysed pollen. 
In fact, acetolysed TSK pollen is even further apart from the rest of the acetolysed pollen than SPT pollen is.
There seem to be chemical differences between TSK and the other two cores that are present in SPT and acetolysed pollen.
One reason for the dissimilarity of SPT TSK pollen is the BSi peak at 1630 which is only present in TSK. []
Still, even acetolysed pollen from TSK is quite different from the other cores according to our PLSR model.
The main difference of TSK pollen to the other cores is that TSK pollen is much younger than the other two cores.
Our TSK samples are from the past 80 years, while MFM covers the YD/Holocene transition and DAL samples are from throughout the Holocene.
The other difference is that TSK is a varved core, while MFM and DAL are not.
Our findings suggest a chemical difference in fossil pollen grains, either caused by diagenesis (varved vs non-varved), age or combination of the two that is persistent even after acetolysis.

<!--
.depth patterns
no clear patterns with depths chemically. the more depths are added, the more inaccurate the LDA became.
for acetolysed pollen samples with the youngest and oldest pollen were used, while SPT had all samples.
-->

.application for palaeoecology
Paragraph on practicality

@@

Further studies with GCMS methods on fossil pollen and several species have to be performed.

.conclusions

>>>>

Our aim with this study was to explore the use of a new type of detector (FPA-MCT) in combination with a FTIR microscope to explore the chemical variations of fossil pollen.
Our goal was also to explore the effects of two differemt sample-processing methods on fossil pollen and compare the spectra we obtained with modern samples.

@@

After visual inspection of the spectra, the chemical differences between modern and fossil pollen were explored with a PLSR model.
The ordination showed that the fossil pollen are clearly separated from the modern pollen (Fig \@ref(fig:ordinations) a).
The regression coefficients reflect our observations from the visual inspection and show that lipid  and protein peaks (1745 and 1500 - 1800 cm^-1^) have disappeared and there are substantial changes in the carbohydrate region (1000 to 1250 cm^-1^) in fossil pollen (Fig \@ref(fig:ordinations) b and c).

ADD COMPARISON TO JARDINE PAPERS and other (e.g. Woutersen)
Comparisons to other studies that looked at acetolysed and fossil samples show similarities between our acetolysed *Pinus* pollen and other taxa.
E.g. the carb region 1000 - 1200 in our Pine is most similar to acetolysed extant poaceae set of @jardineSporopolleninChemistryIts2021, two large peaks at 1040 and 1190/1170.

The fossil poaceae data set (Amazon Fan and Western Amazon) from @jardineSporopolleninChemistryIts2021 on the other looks quite different from our fossil pollen and their extant acetolysed set, with a large peak at 1600 and 1190 peak not visible.
These fossil poaceae set was treated differently than ours and jardines extant poaceae set (combination of heavy liquid separation, Na~4~P~2~O~7~, lignite).

@jardineImpactOxidationSpore2015 showed limited effect on pollen chemistry of cold acetolysis if it was shorter than 10 minutes.
Hot acetolysis (90), which is standard procedure in palynology, had an effect after the first minute. 
The simililarities of the acetolysed extant set to our acetolysed fossil set and dissimilarities to pollen treated with a different combination of extractions (Amazon Western and Fan), suggest that chemical treatment of the pollen has a large effect on the chemistry of pollen, both modern and fossil origin.
This is important to consider going forward for the application of IR methods in palaeoecology, where extraction methods are often dependent on sediment type.
The effects of these extraction methods need to be explored on modern and fossil material side by side to see if there are links to bridge the transition from modern to fossil pollen.

## Differences between fossil pollen

### Anomalies in non acetolysed pollen spectra

@@

The 1090 cm^-1^ peak dissappears after acetolysis from all spectra as does the 1630 cm^-1^ peak in Tiefer See. 
The region 1000 - 1200 cm^-1^ is now dominated in acetolysed fossil pollen by two peaks and its shoulders at 1040 and 1170 cm^-1^.
This 1040 cm^-1^ has high importance for the PLSR model differentiating acetolysed and non acetolysed pollen (Fig. \@ref(fig:ordinations) e and f).
This peak in the non acetolysed samples is most likely biogenic silica, when compared to this biogenic silica spectra from diatoms [@alipourInfraredMicrospectroscopyBionanomaterials2016] and hydrated silica spectra [@SilicaATRIRSpectrum].
During the measurements with the microscope, we applied our extracted pollen as droplets onto ZnSe slides and waited for the water to evaporate.
After the droplets had dried we observed a crystalline layer over the same area that covered all particles and pollen (ref Foto in SM).
From the chemical spectra, it is most likely that that it is dissolved silica that crystallised, although the acetolysis treatment removed this peak, which may be because of the extended washing with several chemicals and water during standard acetolysis.

[FM: would probably have to replicate this with BSi in different concentrations and maybe see if thickness of this BSi crystall has an effect on the spectra]
**I think you need to explain this peak in the context of palaeoecological research a bit- could you talk about the correction factors that might be needed if you were to apply this method everywhere? And that different sediment context might create different baseline spectra so understanding sedimentary context is essential?**

### pollen chemical differences

Acetolysed pollen showed two dominant peaks in the carbohydrate region which are not visible in the modern spectra at 1040 and 1170 - 1190 cm^-1^ or the non acetolysed pollen.
This change in the 1170 - 1190 cm^-1^ region is unique to acetolysized pollen and not visible in non acetolysed pollen and had higher regression coefficients that the surrounding wavenumbers (Fig. \@ref(fig:ordinations)).

The presence of this new peak in the shoulder of the 1170 cm^-1^ peak, which can indicate sporopollenin and carbohydrates, suggest changes in the pollen chemistry.
These changes may be related to removal and or alteration of grain wall carbohydrates (cellulose etc) as well as changes to sporopollenin due to acetolysis. [FM: this would be answered by acetolysis of the fresh grains!!]
In non acetolysed pollen the 1190 peak is not visible in any of the cores, which indicates that it is not due to aging/diageneis, but acetolysis.

Using the second PLSR model the acetolysed pollen is clearly separated from the non acetolysed pollen, with a much higher distance between samples for acetolysed pollen compared to non acetolysed pollen (Fig. \@ref(fig:ordinations) d).
In addition, there is some differentiation between the cores, samples from Tiefer See are more separate from samples of the other two cores, which show more similar scores in the ordination.

This seems to indicate that *Pinus* pollen from DAL and MFM are chemically quite similar, while TSK *Pinus* pollen is different.
In fact, acetolysed TSK pollen is even further apart from the rest of the acetolysed pollen than non acetolysed pollen is.
There seem to be chemical differences between TSK and the other two cores that are present in non acetolysed and acetolysed pollen.
The main difference of TSK pollen to the other cores is that TSK pollen is much younger than the other two cores.
Our TSK samples are from the past 80 years, while MFM covers the YD/Holocene transition and DAL samples are from throughout the Holocene, with the youngest sample several 100 years older than TSK pollen.
We cannot see any evidence that sediment type has an effect on the chemistry of the pollen, both TSK and MFM cores are varved cores, while DAL is not.
This would suggest more difference between varved and non varved pollen if different types of diagenesis had an effect. 
Any difference may be overshadowed by the large difference in age between the samples from these cores. 
Our findings suggest a chemical difference in fossil pollen grains that is persistent even after acetolysis, which may be explained by the age of the pollen.
Further studies on pollen from different sediment types at fixed age/depths may provide a clearer answer on the effects of sediment type and age on pollen chemistry.

@@

@@
